---
title: "Predation_behav_cogcode_for_submission"
output: html_document
date: "2026-05-05"
---


load dataframe
```{r}
library(RCurl)
x <- getURL("https://raw.githubusercontent.com/inmanc2/predation_Gaff_Plat/refs/heads/main/Predation_dataset_foranalysis.csv")
Predation_data <- read.csv(text = x)

```

PCA on scototaxis traits
```{r}

library(dplyr)
library(ggplot2)

# transform relevant variables and omit NAs

df_pca <- Combined_SL_filtered_final %>%
  transmute(
    Predation_Regime,
    Standard_length,
    Sex,
    Population,
    Species,
    Waterway,
    Shyness = log(TimeinBlackThigmo + 0.01),
    Boldness = log(TimeinWhite + 0.01),
    Exploration = log(Distance_moved)
  ) %>%
  na.omit()

```

```{r}

# perform PCA and select relevant transformed variables

pca_mod <- prcomp(
  df_pca %>% select(Shyness, Boldness, Exploration),
  scale. = TRUE
)

# generate PC scores

scores <- as.data.frame(pca_mod$x) %>%
  bind_cols(df_pca %>% select(Predation_Regime, Species, Sex, Population, Standard_length, Waterway))

# generate PC loadings

loadings <- as.data.frame(pca_mod$rotation)
loadings$Trait <- rownames(loadings)

# multiply loadings by scale factor to ease interpretation

scale_factor <- 3
loadings <- loadings %>%
  mutate(
    PC1 = PC1 * scale_factor,
    PC2 = PC2 * scale_factor
  )

```

MANOVA on all behavioral and cognitive traits in dataset
```{r}
library(car)

library(effectsize)

fit <- manova(
 cbind(log(Detour_solve_time_imputed), log(Detour_socialmotivation), log(TimeinBlackThigmo+0.01), log(TimeinWhite+0.01), log(Distance_moved)) ~ Standard_length*Predation_Regime + Sex*Species*Predation_Regime + Waterway + Species + Predation_Regime + Waterway*Predation_Regime,
 data = Combined_SL_filtered_final
)

#type 3 ANOVA on MANOVA fit

Anova(fit, test = "Wilks", type = 3)

eta_squared(fit, partial = TRUE)

```

LDA on all traits

```{r}

library(MASS)

#group based species and predation
Combined_SL_filtered_final$Group1 <- interaction(
  Combined_SL_filtered_final$Species,
  Combined_SL_filtered_final$Predation_Regime,
  sep = "_",
  drop = TRUE
)

traits <- c("Detour_solve_time_imputed",
            "Detour_socialmotivation",
            "TimeinBlackThigmo",
            "TimeinWhite",
            "Distance_moved")

```


```{r}


df_lda_all <- Combined_SL_filtered_final %>%
  dplyr::select(Group1, all_of(traits), Standard_length,
                Waterway, Predation_Regime, Species, Sex, Population) %>%
  na.omit() %>%
  mutate(Group1 = as.factor(Group1))

# add standard length as a covariate and make an object for predictor variables

predictor_vars <- c(traits, "Standard_length")

lda_formula <- as.formula(
  paste("Group1 ~", paste(predictor_vars, collapse = " + "))
)

lda_model_all <- lda(lda_formula, data = df_lda_all)

lda_pred_all <- predict(lda_model_all)

df_lda_all <- df_lda_all %>%
  mutate(
    LD1 = lda_pred_all$x[, 1],
    LD2 = if (ncol(lda_pred_all$x) >= 2) lda_pred_all$x[, 2] else NA
  )


```
ANOVA on LDs
```{r}
anova_ld1 <- aov(LD1 ~ Group1, data = df_lda_all)
summary(anova_ld1)

TukeyHSD(anova_ld1)

anova_ld2 <- aov(LD2 ~ Group1, data = df_lda_all)
summary(anova_ld2)

TukeyHSD(anova_ld2)


```

Set of LMs #1: Trait ~ Predation * Watershed + Predation * Species + Predation * Standard Length + Sex 

```{r}

library(purrr)
library(ggplot2)
library(emmeans)
library(dplyr)


# Prepare data

dat <- Combined_SL_filtered_final %>%
  mutate(
    Species = factor(Species, levels = c("G. affinis", "P. latipinna")),
    Predation_Regime = factor(Predation_Regime, levels = c("LP", "HP")),
    Waterway = factor(Waterway),
    
    Standard_length_c = Standard_length - mean(Standard_length, na.rm = TRUE),

    TimeinBlackThigmo_log = log(TimeinBlackThigmo + 0.01),
    TimeinWhite_log = log(TimeinWhite + 0.01),
    Distance_moved_log = log(Distance_moved+0.01),
    Detour_solve_time_imputed_log = log(Detour_solve_time_imputed+0.01),
    Detour_socialmotivation_log = log(Detour_socialmotivation+0.01)
  )

# Trait definitions

trait_vars <- c(
  "TimeinBlackThigmo_log",
  "TimeinWhite_log",
  "Distance_moved_log",
  "Detour_socialmotivation_log",
  "Detour_solve_time_imputed_log"
)

trait_labels <- c(
  "TimeinBlackThigmo_log"         = "Shyness",
  "TimeinWhite_log"               = "Boldness",
  "Distance_moved_log"            = "Exploration",
  "Detour_socialmotivation_log"   = "Social motivation",
  "Detour_solve_time_imputed_log" = "Cognitive flexibility"
)



library(car)
library(dplyr)
library(purrr)
library(tibble)

final_lm_results <- imap_dfr(
  model_list,
  function(model, trait) {

    a <- car::Anova(model, type = 3) %>%
      as.data.frame() %>%
      rownames_to_column("Effect")

    a %>%
      filter(!Effect %in% c("(Intercept)", "Residuals")) %>%
      transmute(
        Trait = trait,
        Effect = Effect,
        NumDF = Df,
        DenDF = df.residual(model),
        F_value = `F value`,
        p_raw = `Pr(>F)`,
        partial_eta2 =
          (F_value * NumDF) /
          (F_value * NumDF + DenDF)
      )
  }
)

final_lm_results <- final_lm_results %>%
  group_by(Effect) %>%
  mutate(
    p_FDR = p.adjust(p_raw, method = "fdr")
  ) %>%
  ungroup()

final_lm_results %>%
  print(n = Inf, width = Inf)



```
Updated emmeans contrasts across all focal LMs
```{r}
library(emmeans)

# Exploration
emm_exploration <- emmeans(
  model_list[["Distance_moved_log"]],
  ~ Predation_Regime | Species
)

pairs(emm_exploration, adjust = "none")


# Boldness
emm_boldness <- emmeans(
  model_list[["TimeinWhite_log"]],
  ~ Predation_Regime | Species
)

pairs(emm_boldness, adjust = "none")


# Shyness
emm_shyness <- emmeans(
  model_list[["TimeinBlackThigmo_log"]],
  ~ Predation_Regime | Species
)

pairs(emm_shyness, adjust = "none")

library(emmeans)

# Exploration: HP-LP contrasts
# within each Species × Waterway

exploration_waterway <- emmeans(
  model_list[["Distance_moved_log"]],
  ~ Predation_Regime | Species * Waterway
)

exploration_waterway

pairs(
  exploration_waterway,
  adjust = "none"
)


# Boldness: HP-LP contrasts
# within each Species × Waterway


boldness_waterway <- emmeans(
  model_list[["TimeinWhite_log"]],
  ~ Predation_Regime | Species * Waterway
)

boldness_waterway

pairs(
  boldness_waterway,
  adjust = "none"
)


library(emmeans)

shyness_waterway <- emmeans(
  model_list[["TimeinBlackThigmo_log"]],
  ~ Predation_Regime | Species * Waterway
)

shyness_waterway

pairs(
  shyness_waterway,
  adjust = "none"
)


# Estimated relationship between body size and shyness
# separately for each predation regime
shyness_size_trends <- emtrends(
  model_list[["TimeinBlackThigmo_log"]],
  ~ Predation_Regime,
  var = "Standard_length_c"
)

shyness_size_trends

# Test whether the size slopes differ between predation regimes
pairs(shyness_size_trends)


shyness_size_species_model <- lm(
  TimeinBlackThigmo_log ~
    Waterway * Predation_Regime +
    Standard_length_c * Predation_Regime * Species +
    Sex,
  data = model.frame(model_list[["TimeinBlackThigmo_log"]])
)

car::Anova(
  shyness_size_species_model,
  type = 3
)


shyness_size_species_trends <- emtrends(
  shyness_size_species_model,
  ~ Predation_Regime | Species,
  var = "Standard_length_c"
)

shyness_size_species_trends

# LP vs HP difference in size slopes within each species
pairs(shyness_size_species_trends)

library(car)

car::Anova(
  shyness_size_species_model,
  type = 3
)


```

function for estimated marginal means tables
```{r}

trait_labels <- c(
  Detour_socialmotivation_log = "Social latency",
  Detour_solve_time_imputed_log = "Solve latency",
  Distance_moved_log = "Exploration",
  TimeinWhite_log = "Boldness",
  TimeinBlackThigmo_log = "Shyness"
)


# Helper to extract each trait


make_emm_table <- function(model, trait_name) {

  # 1. Overall predation-regime contrast
  overall <- emmeans(
    model,
    ~ Predation_Regime
  ) |>
    pairs(adjust = "none") |>
    as.data.frame() |>
    transmute(
      Response = trait_name,
      Explanatory_variable = "Predation regime",
      Contrast = contrast,
      Estimate = estimate,
      t_ratio = t.ratio,
      p_value = p.value
    )

  # 2. Predation-regime contrast within watershed
  by_waterway <- emmeans(
    model,
    ~ Predation_Regime | Waterway
  ) |>
    pairs(adjust = "none") |>
    as.data.frame() |>
    transmute(
      Response = trait_name,
      Explanatory_variable = "Predation regime × Waterway",
      Contrast = paste0(Waterway, ": ", contrast),
      Estimate = estimate,
      t_ratio = t.ratio,
      p_value = p.value
    )

  # 3. Predation-regime contrast within species
  by_species <- emmeans(
    model,
    ~ Predation_Regime | Species
  ) |>
    pairs(adjust = "none") |>
    as.data.frame() |>
    transmute(
      Response = trait_name,
      Explanatory_variable = "Predation regime × Species",
      Contrast = paste0(Species, ": ", contrast),
      Estimate = estimate,
      t_ratio = t.ratio,
      p_value = p.value
    )

  # 4. Difference in standard-length slopes between predation regimes
  size_trends <- emtrends(
    model,
    ~ Predation_Regime,
    var = "Standard_length_c"
  ) |>
    pairs(adjust = "none") |>
    as.data.frame() |>
    transmute(
      Response = trait_name,
      Explanatory_variable = "Predation regime × Standard length",
      Contrast = "LP - HP (slope difference)",
      Estimate = estimate,
      t_ratio = t.ratio,
      p_value = p.value
    )

  bind_rows(
    overall,
    by_waterway,
    by_species,
    size_trends
  )
}

# Build full supplemental table

supp_emm_table <- imap_dfr(
  model_list,
  ~ make_emm_table(
    model = .x,
    trait_name = unname(trait_labels[.y])
  )
)

supp_emm_table
```



Primary and sensitivity model comparison for LM without Sex interaction and LM with sex interaction
```{r}

library(car)
library(dplyr)
library(tibble)

focal_effects <- c(
  "Predation_Regime",
  "Waterway:Predation_Regime",
  "Predation_Regime:Species",
  "Predation_Regime:Standard_length_c"
)

trait_labels <- c(
  TimeinBlackThigmo_log = "Black thigmotaxis",
  TimeinWhite_log = "Time in white (boldness)",
  Distance_moved_log = "Distance moved (exploration)",
  Detour_socialmotivation_log = "Detour engagement latency",
  Detour_solve_time_imputed_log = "Detour solve latency"
)

all_results <- list()

for (trait in names(model_list)) {

  model <- model_list[[trait]]

  # Standardized residuals
  std_resid <- rstandard(model)

  # Keep observations with |standardized residual| <= 3
  keep <- abs(std_resid) <= 3

  # Exact model-frame data used by the primary model
  mf <- model.frame(model)

  # Refit sensitivity model
  sensitivity_model <- lm(
    formula(model),
    data = mf[keep, ]
  )

  # Primary Type III table
  primary <- car::Anova(model, type = 3) |>
    as.data.frame() |>
    rownames_to_column("Effect") |>
    filter(Effect %in% focal_effects) |>
    select(
      Effect,
      F_primary = `F value`,
      p_primary = `Pr(>F)`
    )

  # Sensitivity Type III table
  sensitivity <- car::Anova(sensitivity_model, type = 3) |>
    as.data.frame() |>
    rownames_to_column("Effect") |>
    filter(Effect %in% focal_effects) |>
    select(
      Effect,
      F_sensitivity = `F value`,
      p_sensitivity = `Pr(>F)`
    )

  # Combine
  result <- merge(
    primary,
    sensitivity,
    by = "Effect"
  )

  result$Trait <- unname(trait_labels[trait])
  result$N_primary <- nobs(model)
  result$N_excluded <- sum(!keep)
  result$N_sensitivity <- sum(keep)

  all_results[[trait]] <- result
}

table_s7 <- do.call(
  rbind,
  all_results
)

row.names(table_s7) <- NULL

table_s7

table_s7_short <- table_s7 %>%
  select(
    Trait,
    Effect,
    N_excluded,
    p_primary,
    p_sensitivity
  )

table_s7_short


```



Diagnostics 


```{r}

library(DHARMa)
library(dplyr)
library(purrr)
library(tibble)


trait_labels <- c(
  TimeinBlackThigmo_log = "Black thigmotaxis",
  TimeinWhite_log = "Time in white (boldness)",
  Distance_moved_log = "Distance moved (exploration)",
  Detour_socialmotivation_log = "Detour engagement latency",
  Detour_solve_time_imputed_log = "Detour solve latency"
)

set.seed(123)

diagnostic_table <- imap_dfr(
  model_list,
  function(model, trait) {


    r <- rstandard(model)


    # DHARMa simulated residuals

    sim <- simulateResiduals(
      fittedModel = model,
      n = 1000,
      plot = FALSE
    )

    uniformity <- testUniformity(sim)
    dispersion <- testDispersion(sim)
    outliers <- testOutliers(sim)

    # --------------------------------
    # Compile results
    # --------------------------------
    tibble(
      Trait = unname(trait_labels[trait]),

      N = nobs(model),

      DHARMa_KS_D =
        unname(uniformity$statistic),

      DHARMa_uniformity_p =
        uniformity$p.value,

      DHARMa_dispersion =
        unname(dispersion$statistic),

      DHARMa_dispersion_p =
        dispersion$p.value,

      DHARMa_outlier_p =
        outliers$p.value,

      Residuals_over_3 =
        sum(abs(r) > 3),

      Percent_over_3 =
        100 * mean(abs(r) > 3),

      Minimum_std_residual =
        min(r),

      Maximum_std_residual =
        max(r),

      Maximum_abs_std_residual =
        max(abs(r))
    )
  }
)

diagnostic_table



diagnostic_table_final <- diagnostic_table %>%
  mutate(
    Extreme_residuals = paste0(
      Residuals_over_3,
      " (",
      round(Percent_over_3, 1),
      "%)"
    )
  ) %>%
  select(
    Trait,
    N,
    DHARMa_uniformity_p,
    DHARMa_dispersion,
    DHARMa_dispersion_p,
    DHARMa_outlier_p,
    Extreme_residuals
  )

diagnostic_table_final

diagnostic_table_final %>%
  print(n = Inf, width = Inf)
  
  diagnostic_table %>%
  select(
    Trait,
    Residuals_over_3,
    Percent_over_3,
    Minimum_std_residual,
    Maximum_std_residual,
    Maximum_abs_std_residual
  ) %>%
  print(n = Inf, width = Inf)

diagnostic_table %>%
  select(
    Trait,
    DHARMa_outlier_p
  ) %>%
  print(n = Inf, width = Inf)


```



Set of LMs #2 (for visualization only, Figures 3-4), includes 3-way interaction: Predation Regime * Watershed * Species
```{r}

model_list2 <- setNames(vector("list", length(trait_vars)), trait_vars)

for (trait in trait_vars) {
  form <- as.formula(
    paste0(
      trait,
      " ~ Waterway * Species *Predation_Regime + ",
      "Standard_length_c * Predation_Regime +",
      "Sex +",
      "(1 | Population)"
    )
  )

  model_list2[[trait]] <- lmer(form, data = dat, REML = FALSE)
}


# Estimated marginal means for model set 2

library(emmeans)

mod2 <- model_list2[["trait"]]

emm_all <- emmeans(
  mod2,
  ~ Species | Waterway * Predation_Regime
)

contrast_all <- contrast(emm_all, method = "pairwise")
summary(contrast_all, infer = TRUE)

```




