---
title: "Predation_behav_cogcode_for_submission"
output: html_document
date: "2026-05-05"
---


load dataframe
```{r}
library(RCurl)
x <- getURL("https://raw.githubusercontent.com/inmanc2/predation_Gaff_Plat/refs/heads/main/Predation_dataset_foranalysis.csv")
Predation_data <- read.csv(text = x)

```

PCA on scototaxis traits
```{r}

library(dplyr)
library(ggplot2)

# transform relevant variables and omit NAs

df_pca <- Combined_SL_filtered_final %>%
  transmute(
    Predation_Regime,
    Standard_length,
    Sex,
    Population,
    Species,
    Waterway,
    Shyness = log(TimeinBlackThigmo + 0.01),
    Boldness = log(TimeinWhite + 0.01),
    Exploration = log(Distance_moved)
  ) %>%
  na.omit()

```

```{r}

# perform PCA and select relevant transformed variables

pca_mod <- prcomp(
  df_pca %>% select(Shyness, Boldness, Exploration),
  scale. = TRUE
)

# generate PC scores

scores <- as.data.frame(pca_mod$x) %>%
  bind_cols(df_pca %>% select(Predation_Regime, Species, Sex, Population, Standard_length, Waterway))

# generate PC loadings

loadings <- as.data.frame(pca_mod$rotation)
loadings$Trait <- rownames(loadings)

# multiply loadings by scale factor to ease interpretation

scale_factor <- 3
loadings <- loadings %>%
  mutate(
    PC1 = PC1 * scale_factor,
    PC2 = PC2 * scale_factor
  )

```

MANOVA on all behavioral and cognitive traits in dataset
```{r}
library(car)

fit <- manova(
 cbind(log(Detour_solve_time_imputed), log(Detour_socialmotivation), log(TimeinBlackThigmo+0.01), log(TimeinWhite+0.01), log(Distance_moved)) ~ Standard_length + Sex + Species*Predation_Regime + Waterway + Species + Predation_Regime + Waterway*Predation_Regime,
 data = Combined_SL_filtered_final
)

#type 3 ANOVA on MANOVA fit

Anova(fit, test = "Wilks", type = 3)

eta_squared(fit, partial = TRUE)

```

LDA on all traits

```{r}

library(MASS)

#group based species and predation
Combined_SL_filtered_final$Group1 <- interaction(
  Combined_SL_filtered_final$Species,
  Combined_SL_filtered_final$Predation_Regime,
  sep = "_",
  drop = TRUE
)

traits <- c("Detour_solve_time_imputed",
            "Detour_socialmotivation",
            "TimeinBlackThigmo",
            "TimeinWhite",
            "Distance_moved")

```


```{r}


df_lda_all <- Combined_SL_filtered_final %>%
  dplyr::select(Group1, all_of(traits), Standard_length,
                Waterway, Predation_Regime, Species, Sex, Population) %>%
  na.omit() %>%
  mutate(Group1 = as.factor(Group1))

# add standard length as a covariate and make an object for predictor variables

predictor_vars <- c(traits, "Standard_length")

lda_formula <- as.formula(
  paste("Group1 ~", paste(predictor_vars, collapse = " + "))
)

lda_model_all <- lda(lda_formula, data = df_lda_all)

lda_pred_all <- predict(lda_model_all)

df_lda_all <- df_lda_all %>%
  mutate(
    LD1 = lda_pred_all$x[, 1],
    LD2 = if (ncol(lda_pred_all$x) >= 2) lda_pred_all$x[, 2] else NA
  )


```
ANOVA on LDs
```{r}
anova_ld1 <- aov(LD1 ~ Group1, data = df_lda_all)
summary(anova_ld1)

TukeyHSD(anova_ld1)

anova_ld2 <- aov(LD2 ~ Group1, data = df_lda_all)
summary(anova_ld2)

TukeyHSD(anova_ld2)


```

Set of LMMs #1: Trait ~ Predation * Watershed + Predation * Species + Predation * Standard Length + Sex + (1 | Population)

```{r}

library(purrr)
library(ggplot2)
library(lmerTest)
library(emmeans)
library(dplyr)


# Prepare data

dat <- Combined_SL_filtered_final %>%
  mutate(
    Species = factor(Species, levels = c("G. affinis", "P. latipinna")),
    Predation_Regime = factor(Predation_Regime, levels = c("LP", "HP")),
    Waterway = factor(Waterway),
    
    Standard_length_c = Standard_length - mean(Standard_length, na.rm = TRUE),

    TimeinBlackThigmo_log = log(TimeinBlackThigmo + 0.01),
    TimeinWhite_log = log(TimeinWhite + 0.01),
    Distance_moved_log = log(Distance_moved+0.01),
    Detour_solve_time_imputed_log = log(Detour_solve_time_imputed+0.01),
    Detour_socialmotivation_log = log(Detour_socialmotivation+0.01)
  )

# Trait definitions

trait_vars <- c(
  "TimeinBlackThigmo_log",
  "TimeinWhite_log",
  "Distance_moved_log",
  "Detour_socialmotivation_log",
  "Detour_solve_time_imputed_log"
)

trait_labels <- c(
  "TimeinBlackThigmo_log"         = "Shyness",
  "TimeinWhite_log"               = "Boldness",
  "Distance_moved_log"            = "Exploration",
  "Detour_socialmotivation_log"   = "Social motivation",
  "Detour_solve_time_imputed_log" = "Cognitive flexibility"
)


# Fit one model per trait

model_list <- setNames(vector("list", length(trait_vars)), trait_vars)

for (trait in trait_vars) {
  form <- as.formula(
    paste0(
      trait,
      " ~ Waterway * Predation_Regime + ",
      "Species * Predation_Regime + ",
      "Standard_length_c * Predation_Regime+",
      "Sex +",
      "(1 | Population)"
    )
  )

  model_list[[trait]] <- lmer(form, data = dat, REML = FALSE)
}



```
Generate ANOVA table of effects from LMMs
```{r}

anova_results <- imap_dfr(model_list, function(model, trait) {
  anova(model, type = 3) %>%
    as.data.frame() %>%
    rownames_to_column("Effect") %>%
    mutate(Trait = trait) %>%
    select(
      Trait,
      Effect,
      NumDF = NumDF,
      DenDF = DenDF,
      F_value = `F value`,
      p_value = `Pr(>F)`
    )
})
 
anova_results_fdr <- anova_results %>%
  group_by(Effect) %>%
  mutate(p_FDR = p.adjust(p_value, method = "fdr")) %>%
  ungroup()


anova_results_fdr


```
Estimated marginal means for set of LMMs #1
```{r}
library(emmeans)

mod <- model_list[["trait"]]

emm_all <- emmeans(
  mod,
  ~ Predation_Regime | Waterway * Species
)

contrast_all <- contrast(emm_all, method = "pairwise")
summary(contrast_all, infer = TRUE)

```

Set of LMMs #2 (for visualization only, Figures 3-4), includes 3-way interaction: Predation Regime * Watershed * Species
```{r}

model_list2 <- setNames(vector("list", length(trait_vars)), trait_vars)

for (trait in trait_vars) {
  form <- as.formula(
    paste0(
      trait,
      " ~ Waterway * Species *Predation_Regime + ",
      "Standard_length_c * Predation_Regime +",
      "Sex +",
      "(1 | Population)"
    )
  )

  model_list2[[trait]] <- lmer(form, data = dat, REML = FALSE)
}


# Estimated marginal means for model set 2

library(emmeans)

mod2 <- model_list2[["trait"]]

emm_all <- emmeans(
  mod2,
  ~ Species | Waterway * Predation_Regime
)

contrast_all <- contrast(emm_all, method = "pairwise")
summary(contrast_all, infer = TRUE)

```




